Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

### 2025-03-02 code/processing-code/processing-code-tickcases.qmd
Part Two Edits

# Nice visualization of data
skimr::skim(rawdata)
```
The first cleaning step removes cases where the STATUS field is either missing or not labeled as "CONFIRMED." This ensures that only officially confirmed cases are included in further analysis. The cleaned dataset is then saved for future reference.

@@

```{r}
## ---- Table --------
cleaned_data %>%
  dplyr::select(DISEASE, AGE, GENDER, COUNTY, STATE, LABDATE1, RESULT1, STATUS) %>%  # Select key columns for display
  knitr::kable(caption = "Cleaned Tick-Bourne Disease Cases (2019-2024)", align = "c") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(2:8, width = "12em") %>%  # Adjust column widths for readability
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Style header row
```
To refine the dataset further, only cases with a confirmed "positive" test result are retained. Any cases with missing, blank, or ambiguous results (such as "inconclusive") are excluded. This step eliminated 15 cases that did not meet the case definition used in Georgia, leaving 95 valid data points. However, upon further review, I noticed that additional positive cases that were not explicitly labeled "positive" were inadvertently removed, prompting a refinement of the filtering criteria.

@@

## ---- display_table_step2 --------
# Display professional table for second cleaning step
kableExtra::kable(cleaned_data2, format = "html", caption = "Final Cleaned Data After Removing Blank RESULT1 Cases") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight first column
  kableExtra::column_spec(2:ncol(cleaned_data2), width = "10em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Header row styling
```
This step provided an easier summary table of some of the most relevant information on these positive cases.

@@

# Display the table with consistent styling
kableExtra::kable(selected_columns_data, format = "html", caption = "Selected Columns from Final Cleaned Data") %>%
  kableExtra::kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(1:ncol(selected_columns_data), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling
```
A second round of filtering ensures that all positive cases are correctly identified while excluding known non-positive values such as "Negative," "Equivocal," "Not detected," and other irrelevant entries. The updated dataset is saved again for consistency.
```{r}
## ---- clean_data_step3--------
# Define values to exclude from RESULT1
excluded_values <- c("", NA, "Negative", "Equivocal", "negative", "Choose One", "Not detected", "Indeterminant/Equivocal")

@@

```
To present the cleaned data effectively, selected key columns—including DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are displayed in a professionally styled table using the kableExtra package. This enhances readability and makes it easier to analyze trends.
```{r}
## ---- display only a few of the columns --------

@@

```
A histogram is created to visualize the distribution of different tick-borne diseases in Georgia between 2019 and 2024. The ggplot2 package is used to generate a clean and informative bar chart, illustrating the most prevalent diseases. 
```{r}
## ---- disease_histogram --------
library(ggplot2)

@@

# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
```
To explore the severity of Rocky Mountain Spotted Fever cases, a box plot of TITER1 values (log-transformed for better visualization) is created. This helps in understanding the spread and variability of test results among confirmed cases. It also is considered relatively analagous to the patterns seen in Alpha-Gal Syndrome, which is a topic of interest for me, but is not explicitly included in this data set as it is currently not a reportable disease in Georgia.
```{r}
# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)

@@

```
The dataset is filtered to determine the five most common tick-borne diseases. A bar chart is created to highlight the case count for each of these diseases, providing an overview of the most significant public health concerns.
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)

@@

```
Lyme Disease is likewise similar to Alpha-gal. So, a specific analysis is conducted for Lyme Disease cases, focusing on test results was conducted. Below ia a count histogram generated to compare positive, negative, and other test outcomes, offering insights into testing patterns and diagnosis rates.

@@

```
To look at trends in disease burden as well as "hot-spot" counties, I generated the stacked bar graph below.
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)
library(RColorBrewer)

@@

```
Age distribution might be of interest in the analysis, so this was checked as part of this initial data cleaning. This output shows 0 rows of the cleaned data to be missing ages. As an exploratory analysis, I looked at disease count by age group from my cleaned data set. 
```{r}
missing_age_rows <- cleaned_data3 %>%
  filter(is.na(AGE))

@@

```
Another thing that might be used in my analysis to answer the question of how data completeness evolved over time is cleaning the data to extract data onset information.

By distinguishing between Blank and Text in the STATUS column over time, this plot helps pinpoint periods with high missingness. If there are spikes in Blank values at certain times, it could indicate data entry issues, system errors, or gaps in reporting that need to be addressed before deeper analysis.
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

@@

```
Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time? One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.

>>>>

# Nice visualization of data
skimr::skim(rawdata)
```

The first cleaning step removes cases where the STATUS field is either missing or not labeled as "CONFIRMED." This ensures that only officially confirmed cases are included in further analysis. The cleaned dataset is then saved for future reference.

@@

```{r}
## ---- Table --------
cleaned_data %>%
  dplyr::select(DISEASE, AGE, GENDER, COUNTY, STATE, LABDATE1, RESULT1, STATUS) %>%  # Select key columns for display
  knitr::kable(caption = "Cleaned Tick-Bourne Disease Cases (2019-2024)", align = "c") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(2:8, width = "12em") %>%  # Adjust column widths for readability
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Style header row
```

To refine the dataset further, only cases with a confirmed "positive" test result are retained. Any cases with missing, blank, or ambiguous results (such as "inconclusive") are excluded. This step eliminated 15 cases that did not meet the case definition used in Georgia, leaving 95 valid data points. However, upon further review, I noticed that additional positive cases that were not explicitly labeled "positive" were inadvertently removed, prompting a refinement of the filtering criteria.

@@

## ---- display_table_step2 --------
# Display professional table for second cleaning step
kableExtra::kable(cleaned_data2, format = "html", caption = "Final Cleaned Data After Removing Blank RESULT1 Cases") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight first column
  kableExtra::column_spec(2:ncol(cleaned_data2), width = "10em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Header row styling
```

This step provided an easier summary table of some of the most relevant information on these positive cases.

@@

# Display the table with consistent styling
kableExtra::kable(selected_columns_data, format = "html", caption = "Selected Columns from Final Cleaned Data") %>%
  kableExtra::kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(1:ncol(selected_columns_data), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling
```

A second round of filtering ensures that all positive cases are correctly identified while excluding known non-positive values such as "Negative," "Equivocal," "Not detected," and other irrelevant entries. The updated dataset is saved again for consistency.

```{r}
## ---- clean_data_step3--------
# Define values to exclude from RESULT1
excluded_values <- c("", NA, "Negative", "Equivocal", "negative", "Choose One", "Not detected", "Indeterminant/Equivocal")

@@

To present the cleaned data effectively, selected key columns—including DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are displayed in a professionally styled table using the kableExtra package. This enhances readability and makes it easier to analyze trends.

@@

A histogram is created to visualize the distribution of different tick-borne diseases in Georgia between 2019 and 2024. The ggplot2 package is used to generate a clean and informative bar chart, illustrating the most prevalent diseases.

```{r}
## ---- disease_histogram --------
library(ggplot2)

@@

# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
```

To explore the severity of Rocky Mountain Spotted Fever cases, a box plot of TITER1 values (log-transformed for better visualization) is created. This helps in understanding the spread and variability of test results among confirmed cases. It also is considered relatively analagous to the patterns seen in Alpha-Gal Syndrome, which is a topic of interest for me, but is not explicitly included in this data set as it is currently not a reportable disease in Georgia.

```{r}
# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)

@@

The dataset is filtered to determine the five most common tick-borne diseases. A bar chart is created to highlight the case count for each of these diseases, providing an overview of the most significant public health concerns.

@@

Lyme Disease is likewise similar to Alpha-gal. So, a specific analysis is conducted for Lyme Disease cases, focusing on test results was conducted. Below ia a count histogram generated to compare positive, negative, and other test outcomes, offering insights into testing patterns and diagnosis rates.

@@

To look at trends in disease burden as well as "hot-spot" counties, I generated the stacked bar graph below.

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)
library(RColorBrewer)

@@

Age distribution might be of interest in the analysis, so this was checked as part of this initial data cleaning. This output shows 0 rows of the cleaned data to be missing ages. As an exploratory analysis, I looked at disease count by age group from my cleaned data set.

```{r}
missing_age_rows <- cleaned_data3 %>%
  filter(is.na(AGE))

@@

Another thing that might be used in my analysis to answer the question of how data completeness evolved over time is cleaning the data to extract data onset information.

By distinguishing between Blank and Text in the STATUS column over time, this plot helps pinpoint periods with high missingness. If there are spikes in Blank values at certain times, it could indicate data entry issues, system errors, or gaps in reporting that need to be addressed before deeper analysis.

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

@@

Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time? One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.

### 2025-03-02 products/manuscript/Manuscript.qmd
Part Two Edits

# Generate correct file paths
img_paths <- c(
  here::here("results", "figures", "disease_county_plot.png"),
  here::here("results", "figures", "top5_disease_county_plot.png")
)

>>>>

# Generate correct file paths
img_paths <- c(
  here::here("results", "figures", "disease_county_plot.png"),
  here::here("results", "figures", "top5_diseases_stacked_plot"),
  here::here("results", "figures", "top5_disease_county_plot.png")
)
